---
title: "assignment1"
author: "nihat uzunalioglu - 2660298, emiel kempen - 2640580, saurabh jain - 2666959"
date: "2/26/2020"
output: pdf_document
---

```{r setup, set.seed(43), include=FALSE}
knitr::opts_chunk$set(fig.height = 2.5)
```

## Exercise 1
```{r, echo=FALSE}
bread = read.table(file = 'data/bread.txt', header = TRUE)
```
a) The 18 slices came from a single loaf, but were randomized to the 6 combinations of conditions. Present an R-code for this randomization process.
```{r, echo=TRUE}
# The randomization process for 18 slices
# Take hours column from the data
hours = as.vector(as.matrix(bread$hours))
# Create environment column
environment = rep(c('cold', 'intermediate', 'warm'), each = 6)
# Create humidity column
humidity = rep(c('dry', 'wet'), each = 9)
# Converting to data frame
data.frame(cbind(hours, environment, humidity))
```

b) Make two boxplots of hours versus the two factors and two interaction plots (keeping the two factors fixed in turn).
```{r, echo=FALSE}
# Divide into two columns
par(mfrow=c(1, 2))
# Boxplot vs interaction plot in hours~environment
boxplot(hours~environment, data = bread)
interaction.plot(environment, humidity, hours, fixed = TRUE)
# Boxplot vs interaction plot in hours~humidity
boxplot(hours~humidity, data = bread)
interaction.plot(humidity, environment, hours, fixed = TRUE)
```

c) Perform an analysis of variance to test for effect of the factors temperature, humidity, and their interaction. Describe the interaction effect in words.

```{r, echo=FALSE}
# Creating linear model and ANOVA test
breadaov = lm(hours~environment*humidity, data = bread); anova(breadaov)
p_interaction = anova(breadaov)$Pr[3]
```
- The p-value for testing for $H_{0}$:$\gamma_{i,j}$ = 0 for all i, j is `r p_interaction`. Therefore, we reject the null hypothesis $H_{0}$ which means the interaction between environment and humidity is significant for this dataset.