---
title: "assignment3"
author: "nihat uzunalioglu - 2660298, emiel kempen - 2640580, saurabh jain - 2666959"
date: "3/15/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height = 4)
options(show.signif.stars=FALSE, digits = 3)
```

## Exercise 1

```{r, echo=FALSE}
fruitflies = read.table('data/fruitflies.txt', header = TRUE)
```

```{r, echo=TRUE}
# Add loglongevity to the data-frame
# Use it as a response variable (Y)
fruitflies$loglongevity = log(fruitflies$longevity)
```

#### Section A
```{r, echo=FALSE}
# par(mfrow=c(2,2))
boxplot(loglongevity~activity,data=fruitflies)

# Plotting hist & qq-plot of each activity separately
log_low = fruitflies[fruitflies$activity == 'low',]$loglongevity
# hist(log_low)
qqnorm(log_low, main = 'low activity'); qqline(log_low)

log_iso = fruitflies[fruitflies$activity == 'isolated',]$loglongevity
# hist(log_iso)
qqnorm(log_iso, main = 'isolated activity'); qqline(log_iso)

log_high = fruitflies[fruitflies$activity == 'high',]$loglongevity
# hist(log_high)
qqnorm(log_high, main = 'high activity'); qqline(log_high)

# # Interaction Plots
# attach(fruitflies)
# par(mfrow=c(1,1))
# interaction.plot(thorax, activity, loglongevity)
```

- From the boxplot, we can observe medians are not the same among activities and there is also an outlier both in high and low activity. From QQ-plots, we observed normality on low activity with 2 outliers, whereas isolated and high activities don't have normality.

```{r, echo=FALSE}
# Create model
fruitflies$activity = as.factor(fruitflies$activity)
lm_act=lm(loglongevity~activity,data=fruitflies)
# Take residuals
res = residuals(lm_act)

# Shapiro test
pfromshapiro = shapiro.test(res)[2]$p.value
# ANOVA test
anova(lm_act)
# Summary Coefficients
summary(lm_act)[4]
```

- We reject the null hypothesis of ANOVA test which means there is significant difference among groups and also we can observe the same outcome from the summary coefficient table. According to the estimations of longevity, it increases more when the sexual activity is isolated, then it follows as low and high. To find out the $\alpha$s;
  - $\alpha_{high}$ = $\mu_{1}$ = 3.602,
  - $\alpha_{isolated}$ = $\mu_{2}-\mu_{1}$ = 0.517 ==> $\mu_{2}$ = `r 3.602 + 0.517`,
  - $\alpha_{low}$ = $\mu_{3}-\mu_{1}$ = 0.398 ==> $\mu_{3}$ = `r 3.602 + 0.398`

```{r, echo=FALSE}
par(mfrow=c(1, 2))
# QQ-plot
qqnorm(res)
# Scatter plot
plot(fitted(lm_act), res)
```

- QQ plot does not provide good results with regards to the normality. Shapiro - Wilk test also supplies the same outcome with `r pfromshapiro` value since we rejected null hypothesis $H_{0}$ (because it is lower than 0.05) which means residuals are not normally distributed and that is a sign which tells something wrong about our model, even though, we created our model with log of longevity (a transformation to longevity for the sake of the model beforehand).

#### Section B

```{r, echo=FALSE}
# Create model
thorax_avg = mean(fruitflies$thorax); thorax_avg
fruitflies$thorax = as.factor(fruitflies$thorax)
lm_act2=lm(loglongevity~thorax+activity,data=fruitflies)
# Take residuals
res2 = residuals(lm_act2)
# Shapiro test
pfromshapiro2 = shapiro.test(res2)[2]$p.value
# ANOVA test
anova(lm_act2)
# Summary Coefficients
summary(lm_act2)[4]
```

- According to the two-way ANOVA results, we observe that thorax ($\alpha_{i}$) and activity ($\beta_{j}$) have both main effects on longevity in the additive model since we rejected the null hypothesis.
- From the summary coefficients, we observe that all activity factors increase longevity since all of them are positive.

- We obtain thorax average as `r thorax_avg`

```{r, echo=FALSE}
fruit=data.frame(thorax=fruitflies$thorax,
                 activity=fruitflies$activity,
                 longevity=fruitflies$longevity,
                 loglongevity=fruitflies$loglongevity)
# Summary with sum contrast
fruit$thorax = as.factor(fruit$thorax)
fruit$activity = as.factor(fruit$activity)
contrasts(fruit$thorax) = contr.sum
contrasts(fruit$activity) = contr.sum
# Create model
fruit_contr_sum = lm(loglongevity~thorax+activity, data = fruit)
anova(fruit_contr_sum)
# Summary
summary(fruit_contr_sum)[4]
```

- As contr.sum equals to zero, we calculated activity3 (high) as -(-0.2758 + 0.1865) = `r -(-0.2758 + 0.1865)`. So, to have the estimates for flies with average thorax, we move on with the values thorax9 = 0.3121 (for the average thorax length), and all activity factors.
  - $Y_{isolated, thorax9}$ = 3.8026 + 0.3121 - 0.2758 = `r 3.8026 + 0.3121 - 0.2758`,
  - $Y_{low, thorax9}$ = 3.8026 + 0.3121 + 0.1865 = `r 3.8026 + 0.3121 + 0.1865`,
  - $Y_{high, thorax9}$ = 3.8026 + 0.3121 + `r -(-0.2758 + 0.1865)` = `r 3.8026 + 0.3121 - (-0.2758 + 0.1865)`

#### Section C

```{r, echo=FALSE}
# Plotting
fruitflies$thorax = as.numeric(fruitflies$thorax)
plot(loglongevity~thorax, data=fruitflies[fruitflies$activity == 'high',], col ='red')
points(loglongevity~thorax, data=fruitflies[fruitflies$activity == 'low',], col ='blue')
points(loglongevity~thorax, data=fruitflies[fruitflies$activity == 'isolated',], col ='green')
legend('topleft', legend = c('isolated', 'high', 'low'), fill = c('green', 'red', 'blue'))
```

- There is an increment in longevity as thorax length increases. Moreover, flies which were in isolated sexual activity seem to have the longer than the others. Additionally, it follows as low and isolated activity factors, respectively. But to be sure, we also obtain results from ANOVA additive model.

```{r, echo=FALSE}
# Testing
# Creating model
lm_tho = lm(loglongevity~thorax+activity, data = fruitflies)
# ANOVA test
anova(lm_tho)
# Summary Coefficients
summary(lm_tho)[4]
```

- According to the ANOVA test results, we reject null hypothesis for both activity and thorax which leads to the fact that they have significant effect upon longevity. Whereas we can't say we reject null hypothesis for the interaction between activity and thorax.
  - $\alpha_1$ (high) = $\mu_{1}$ = 3.0762,
  - $\alpha_2$ (isolated) = $\mu_{2}-\mu_{1}$ = 3.0762 + 0.4120 = `r 3.0762 + 0.4120`,
  - $\alpha_3$ (low) = $\mu_{2}-\mu_{1}$ = 3.0762 + 0.2871 = `r 3.0762 + 0.2871`,
  - When we look the the results the highest effect is supplied by low activity, then isolated and lastly high sexual condition.

#### Section D

- We would prefer without thorax parameter since there is no real interaction between activity and thorax. Moreover, as in the beginning of the question, it says that experimenters randomly chose the sexual activity upon flies will going to experience and most importantly, thorax is considered to be added later on which does not seem a reliable factor for this testing.

#### Section E

```{r, echo=FALSE}
par(mfrow=c(1,2))
qqnorm(residuals(fruit_contr_sum)); qqline(residuals(fruit_contr_sum))
plot(fitted(fruit_contr_sum), residuals(fruit_contr_sum))
```

- The normality seems doubtful in QQ-plot and we did not observe heteroscedasticity in the scatter plot as they mostly are accumulated on the right side of the plot.

```{r, echo=FALSE}
library(lmtest)
bptest(fruit_contr_sum)
```

- We also don't reject the null hypothesis of Breusch-Pagan test which is the error variances are all equal and that is a proof of non-heteroscedasticity among the model.

#### Section F

```{r, echo=FALSE}
fruit$activity = as.factor(fruit$activity)
act_fact_mod = lm(fruit$longevity~fruit$thorax+fruit$activity)
drop1(act_fact_mod, type='F')
```

```{r, echo=FALSE}
par(mfrow=c(1, 2))
qqnorm(residuals(act_fact_mod), main = 'number of days'); qqline(residuals(act_fact_mod))
qqnorm(residuals(lm_act2), main = 'log of number of days'); qqline(residuals(lm_act2))

plot(fitted(act_fact_mod), residuals(act_fact_mod), main = 'number of days')
plot(fitted(lm_act2), residuals(lm_act2), main = 'log of number of days')
```

- Normality is doubtful for both of the QQ-plots. But residuals of the number of days seem to be more normal distributed than the log of them.

```{r, echo=FALSE}
bptest(act_fact_mod); bptest(lm_act2)
```

- According to the Breush-Pagan tests for both numerical (first test) and log of longevity values (second test), we don't reject, and therefore, confirm non-heteroscedasticities for both of them. Since we have better QQ-plot and greater p-value from Breush-Pagan test, we conclude as the real numerical values for longevity is better than logarithmic for the model.


## Exercise 2

```{r, echo=FALSE}
psi = read.table(file = 'data/psi.txt', header=TRUE)
```

a)

First, we load the data and check the normality of the GPA data.

```{r, echo=FALSE}
qqnorm(psi$gpa, main="Q-Q Plot: All Students' GPA")
qqline(psi$gpa)
```

From the QQ-plot the normality can not be assumed from the plot, so we perform a Shapiro Test to make sure. The test outputs a p-value of `r shapiro.test(psi$gpa)[2]`, which means the GPA data is normally distributed.

Then, we divide the dataset into two groups based on whether they received psi, to be able to study the data and give summaries.

```{r, echo=TRUE}
no_psi = psi[which(psi$psi == "0"),]
yes_psi = psi[which(psi$psi == "1"),]
```

The splitted data can then be studied from the following boxplot.

```{r, echo=FALSE}
boxplot(psi$gpa, yes_psi$gpa, no_psi$gpa, main="GPA Boxplots", names=c("All", "PSI", "No PSI"));
```

We can see that the mean GPA of students with no psi is lower compared to the mean GPA of students with psi, but the latter has a higher variance.

```{r, echo=FALSE}
par(mfrow=c(1, 2))
hist(no_psi$gpa, breaks = c(2.0, 2.5, 3.0, 3.5, 4.0), xlab = 'GPA', ylab = 'N of Students', main = 'Students that did not receive psi')
hist(yes_psi$gpa, breaks = c(2.0, 2.5, 3.0, 3.5, 4.0), xlab = 'GPA', ylab = 'N of Students', main = 'Students that did receive psi')
```

From the boxplots it clearly shows that students that did receive psi are more lenient towards the higher end of the GPAs, for students without psi the data is more concentrated with GPAs around the 2.5 - 3.0 mark.

b)

Fitting a logistic regression model.

```{r}
psiglm=glm(passed~psi+gpa,data=psi,family=binomial)
summary(psiglm)$coefficients

plot(psiglm, which=2:2)
```

From the summary follows that the p-value for receiving psi equals `r summary(psiglm)$coefficients[12]` so we reject the $H_0$, concluding that receiving psi does work. Also, the parameter estimate of psi of $2.34$ shows that that if psi is received, the probability of passing is higher.
From the graph it shows that the residuals seem quite normal, which shows that our model works well. It also show that points 24, 14 and 32 are outliers in the data set.

c)

From the summary of the previous question follow the following probabilities:

A student with a GPA equal to 3 who receives psi

- $-11.602 + 2.338 + (3.063 * 3.0) = -0.075$
- Probability: $\frac{1}{1+e^{-(-0.075)}} =$ `r 1/(1+exp(0.075))`

A student with a GPA equal to 3 who does not receives psi

- $-11.602 + (3.063 * 3.0) = -2.408$
- Probability: $\frac{1}{1 + e^{-(-2.408)}} =$ `r 1/(1+exp(2.408))`

The probability of a student that receives psi passing the assigment is higher than the student with a same GPA who did not receive psi.

d)

From the same summary follow the following probabilities.

A student that received psi:

- $-11.602 + 2.34 = -9.26$
- Probability: $\frac{1}{1 + e^{-(-9.26)}} =$ `r 1/(1 + exp(9.26))`

A student that did not receive psi:

- Probability: $\frac{1}{1 + e^{-(-11.602)}} =$ `r 1/(1+exp(11.602))`

The relative change in odds is as follows:

- `r 1/(1 + exp(9.26))` $-$ `r 1/(1+exp(11.602))`  $=$ `r 1/(1 + exp(9.26)) - 1/(1+exp(11.602))`

This number means that a student's odds increase by `r 1/(1 + exp(9.26)) - 1/(1+exp(11.602))` to pass the assigment when receiving psi. This relative change is not dependent on GPA, because if we leave GPA out of the GLM function, the parameter estimate for psi - the number that is used in this calculation - does not change.

e)

```{r, echo=TRUE, results='hide'}
x=matrix(c(3,15,8,6),2,2); fisher.test(x)
```

In the matrix x, the numbers 15 and 6 equal the number of students that did not show improvements while not receiving and receiving psi respectively. The conclusion is that the $H_0$ - that states that the probability of passing is the same for the two groups - is rejected due to the p-value of `r fisher.test(x)[1]`.

f)

Fisher's exact test is meant to compute an exact p-value for 2x2 tables. The test is mainly used for small sample sizes, which is also the case in our example. This, combined with the way the experiment was conducted, makes us believe that this approach is not wrong.

g)

The last approach could make use of Fisher's exact test, which produces an exact p-value, rather than an approximation that only becomes exact as the sample size grows to infinity, as with many other statistical tests, the one used in part b) of this question included.

However, Fisher's test, in contrast to logistic regression, can only be used for relatively small sample sizes (i.e., small counts in the 2x2 table). Therefore it was not possible to conduct this test on the data acquired by the experiment with the first approach of this question.


## Exercise 3

```{r}
africa = read.table(file = 'data/africa.txt', header = TRUE)
```
a)
We need to treat explanatory variable `pollib` as a factor variable as it represents three different categories and it is not a numerical variable.
```{r}
africa$pollib = as.factor(africa$pollib)
glm_milt=glm(miltcoup~oligarchy+pollib+parties+pctvote+popn+size+numelec+numregim,family=poisson,data=africa) 
summary(glm_milt)[[12]]
plot(glm_milt, which=2)
shapiro.test(residuals(glm_milt))[[2]]
```
`oligarchy`, `parties`, and `pollib` variables are significant for the model as the p-values are less than 5%. As per the plot, `Liberia`, `Comoros`, and `Sudan` are the outliers in the data. Also, we reject the $H_0$ hypothesis of 'Shapiro test' and conclude the residuals of the model are not normally distributed, meaning that the model does not work well on the given dataset. 

b)
In 'Step down' method, we start with all the variables and then reduce the number of variables in our model based on the p-value.
```{r}
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote+popn+size+numelec+numregim,family=poisson,data=africa))[[12]]
```
As the variable `numelec` has the highest p-value and it is > 0.05, we discard it for the next iteration.
```{r}
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote+popn+size+numregim,family=poisson,data=africa))[[12]]
```
In this iteration, variable `numregim` have the p-value > 0.05, thus we discard it for the next iteration.
```{r}
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote+popn+size,family=poisson,data=africa))[[12]]
```
In this iteration, variable `size` have the p-value > 0.05, thus we discard it for the next iteration.
```{r}
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote+popn,family=poisson,data=africa))[[12]]
```
In this iteration, variable `popn` have the p-value > 0.05, thus we discard it for the next iteration.
```{r}
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote,family=poisson,data=africa))[[12]]
```
In this iteration, variable `pctvote` have the p-value > 0.05, thus we discard it for the next iteration.

```{r}
#Final Model
model_poisson = glm(miltcoup~oligarchy+pollib+parties,family=poisson,data=africa)
summary(model_poisson)[[12]]
par(mfrow=c(1, 2))
plot(model_poisson, which = 1:2)

```
In the final model obtained using 'Step Down' approach, `oligarchy`, `parties`, and `pollib2` turned out to be most important variables with p-values less than 5%. Also, as per the QQ plot, distribution is evidently normaly distributed with few outliers.
